
```{r setup, include=FALSE}
library(tidyverse)
library(foreign)
library(cowplot)
library(lubridate)
library(rstan)
library(sn)
library(openxlsx)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
knitr::opts_knit$set(root.dir = "~/OneDrive - London School of Hygiene and Tropical Medicine/SARS-CoV-2 serosurveys/Mixture models/Analysis Nature Comms Revision")
```

**Data (OD ratios)**

1. Pre-symptomatic PCR +ve cases, n=14

2. Symptomatic PCR +ve cases, n=55

3. Asymptomatic PCR +ve cases, n=92

4. Pre-covid samples (2018) from adults in Kilifi, n=200

5. Pre-covid samples (2018) from children in Kilifi, n=200

6. Pre-covid samples from Kenyan Blood Transfusion Service (KBTS), n=500

7. KBTS samples period 1 (30 April and 19 June), n=3,362

8. KBTS samples period 2 (20 June and 19 August), n=2,837

9. KBTS samples period 3 (20 August and 30 September), n=3,723
   
9. ANC clinic in Nairobi, n=196

10. ANC clinic in Kilifi (Batch 1, Oct), n=264

11. ANC clinic in Kilifi (Batch 2, Dec), n=155

12. Healthcare workers in Kilfi, n=200

13. Healthcare workers in Nairobi, n=183

14. Healthcare workers in Busia, n=301

15. Truck drivers in Malaba, n=364

16. Truck drivers in Magarini, n=101

17. Truck drivers in Busia, n=365


```{r input data, include=FALSE}

pcrpos <- 
  read.dta("data/pcr_positives_24022021.dta")

knbts_pos_neg_data <- 
  read.csv("data/combined_july_october_sen_spec.csv")

knbts_unknowns_data <- 
  read.csv("data/clean_individual_level_data_2021-01-10.csv")

kilifi_anc_data_oct <- 
  read.csv("data/Krammer_ANC_KCH_Kilifi_30102020.csv")

kilifi_anc_data_dec <-
  read.csv("data/kch_anc_03dec2020.csv")

nairobi_anc_data <- 
  read.csv("data/knh_all_data_Kramer_23102020_updated_ANConly.csv")

hcw_data <-
  read.csv("data/seeck_hcw_analytic_dataset.csv") 
  
trucker_data <-
  read.csv("data/trucker_serosurveillance_mixture_models.csv")

pre_bts <- 
  knbts_pos_neg_data %>%
  filter(sampletype == "Pre-COVID BTS samples, 2018") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

pre_ad_kil <- 
  knbts_pos_neg_data %>%
  filter(sampletype == "Pre-COVID Kilifi adults, 2018") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

pre_ch_kil <- 
  knbts_pos_neg_data %>%
  filter(sampletype == "Pre-COVID Kilifi children, 2018") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

pre_other <- 
  knbts_pos_neg_data %>%
  filter(sampletype == "Endemic HCoV conv plasma" | sampletype == "NIBSC panel (negative)" ) %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

asymp <- 
  knbts_pos_neg_data %>%
  filter(sampletype == "asymptomatic") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

presymp <- 
  knbts_pos_neg_data %>%
  filter(sampletype == "pre-symptomatic") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

symp <- 
  knbts_pos_neg_data %>%
  filter(sampletype == "symptomatic") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

kbts_central <- 
  knbts_unknowns_data %>%
  filter(period3 == 3) %>%
  filter(province == "central") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

kbts_coast <- 
  knbts_unknowns_data %>%
  filter(period3 == 3) %>%
  filter(province == "coast_other") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

kbts_eastern_neastern <- 
  knbts_unknowns_data %>%
  filter(period3 == 3) %>%
  filter(province == "eastern_neastern") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

kbts_mombasa <- 
  knbts_unknowns_data %>%
  filter(period3 == 3) %>%
  filter(province == "coast_mombasa") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

kbts_nairobi <- 
  knbts_unknowns_data %>%
  filter(period3 == 3) %>%
  filter(province == "nairobi") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

kbts_nyanza <- 
  knbts_unknowns_data %>%
  filter(period3 == 3) %>%
  filter(province == "nyanza") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

kbts_rift_valley <- 
  knbts_unknowns_data %>%
  filter(period3 == 3) %>%
  filter(province == "rift valley") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

kbts_western <- 
  knbts_unknowns_data %>%
  filter(period3 == 3) %>%
  filter(province == "western") %>%
  pull("spikeratio") %>%
  na.omit() %>%
  log2()

anc_kil_oct <- 
  kilifi_anc_data_oct %>%
  pull("KRSpikeODratio") %>%
  na.omit() %>%
  log2()

anc_kil_dec <-
  kilifi_anc_data_dec %>%
  pull("spikeODratio") %>%
  na.omit() %>%
  log2()

anc_nairobi <- 
  nairobi_anc_data %>%
  pull("od_ratio") %>%
  na.omit() %>%
  log2() 

hcw_bus <-
  hcw_data %>%
  filter(site == "BUS") %>%
  pull("od_ratio") %>%
  na.omit() %>%
  log2()

hcw_kil <-
  hcw_data %>%
  filter(site == "KLF") %>%
  pull("od_ratio") %>%
  na.omit() %>%
  log2()

hcw_nairobi <-
  hcw_data %>%
  filter(site == "NBI") %>%
  pull("od_ratio") %>%
  na.omit() %>%
  log2()

tru_bus <- 
  trucker_data %>%
  filter(site == "BUS") %>%
  pull("spikeODratio") %>%
  na.omit() %>%
  log2()

tru_mag <- 
  trucker_data %>%
  filter(site == "KLF") %>%
  pull("spikeODratio") %>%
  na.omit() %>%
  log2()

tru_mal <- 
  trucker_data %>%
  filter(site == "MAL") %>%
  pull("spikeODratio") %>%
  na.omit() %>%
  log2()

world_covid <-
  read.csv("data/coronavirus_cases_world.csv") 

all_surv <-
  list(
    "Central" = kbts_central,
    "Coast" = kbts_coast,
    "Eastern/N.Eastern" = kbts_eastern_neastern,
    "Mombasa" = kbts_mombasa,
    "Nairobi" = kbts_nairobi,
    "Nyanza" = kbts_nyanza,
    "Rift Valley" = kbts_rift_valley,
    "Western" = kbts_western,
    "ANC Kilifi Oct" = anc_kil_oct,
    "ANC Kilifi Dec" = anc_kil_dec,
    "ANC Nairobi" = anc_nairobi,
    "HCW Busia" = hcw_bus,
    "HCW Kilifi" = hcw_kil,
    "HCW Nairobi" = hcw_nairobi,
    "TRUCKERS Busia" = tru_bus,
    "TRUCKERS Magarini" = tru_mag,
    "TRUCKERS Malaba" = tru_mal)

validation <- 
  list(
    "PCR+ asymp" = asymp,
    "PCR+ symp" = symp,
    "Pre COVID" = c(pre_bts, pre_ad_kil, pre_ch_kil, pre_other))

``` 

```{r output data, include=FALSE}

all_data <- c(validation, all_surv)

log2odr_all_surveys <- matrix(0, length(all_data$`Pre COVID`), length(all_data))

for(i in seq_along(all_data)){
    length(all_data[[i]]) <- length(validation$`Pre COVID`)
    log2odr_all_surveys[,i] <- all_data[[i]]
}

colnames(log2odr_all_surveys) <- names(all_data)  
log2odr_all_surveys <- as.data.frame(log2odr_all_surveys)
write.xlsx(log2odr_all_surveys, 
           colNames = T,
           file = "output/log2odr_all_surveys.xlsx")
```


**Summary stats for log2 OD ratio pre-COVID and PCR positive samples**

```{r}
sd(validation$`Pre COVID`)
mean(validation$`Pre COVID`)
mean(validation$`PCR+ symp`)
mean(c(validation$`PCR+ asymp`, validation$`PCR+ symp`))
t.test(validation$`PCR+ asymp`, validation$`PCR+ symp`)
sd(c(validation$`PCR+ asymp`, validation$`PCR+ symp`))
```


**Median time between PCR test and blood sample in +ve cases**

```{r}
pcrpos %>%
rename(sampleid = record_id) %>%
inner_join(knbts_pos_neg_data, by = "sampleid") %>%
mutate(diff = blood_sample_date - date_pos_pcr) %>%
summarise(
  mean(diff, na.rm = T),
  min(diff, na.rm = T),
  quantile(diff, c(0.25, 0.5, 0.75), na.rm = T))  

```

**Time between PCR test and blood sample in +ve cases with OD ratio < 2**

```{r}
pcrpos %>%
rename(sampleid = record_id) %>%
inner_join(knbts_pos_neg_data, by = "sampleid") %>%
mutate(diff = blood_sample_date - date_pos_pcr) %>%
filter(KRSpikeOdratio < 2)  %>%
select(KRSpikeOdratio, diff)  
  
```

```{stan, output.var="mixmod1", include=F, cache=T}

data {
  int N;
  vector[N] y;
  real alpha_sd;
  real p_ub;
}

parameters {
  real theta;
  real<lower=0> delta;
  real log_omega;
  real alpha;               
  real<lower=0,upper=p_ub> p;
}

transformed parameters {
  real xi;
  real nu = 0.42;
  real<lower=0> omega;
  real theta_delta;
  omega = exp(log_omega);
  theta_delta = theta + delta;
  xi = theta + delta - sqrt(2/3.142) * omega * alpha/sqrt(1 + alpha^2);
}

model {

  // priors
  theta ~ normal(0, 10);
  delta ~ normal(0, 1.83); 
  log_omega ~ normal(0, 10);
  alpha ~ normal(0, alpha_sd);
  
  //likelihood
  for (n in 1:N) {
    target += log_sum_exp(log(p)
                           + skew_normal_lpdf(y[n] | xi, omega, alpha),
                          log1m(p)
                          + normal_lpdf(y[n] | theta, nu));
  }
}  
  
generated quantities{
    vector[10] cut = [-2, -1.5, -1, -0.5, 0, 0.5, 1, 1.5, 2, 5]';
    vector[10] sp;
    vector[10] se;
    for (i in 1:10) {
      sp[i] = exp(normal_lcdf(cut[i] | theta, nu));
      se[i] = exp(skew_normal_lccdf(cut[i] | xi, omega, alpha));
  }
}

```


```{r, eval=F, include=F}
 traceplot(fit, 
           pars = c("theta", "xi", "omega", "delta", "alpha", "p"), 
           inc_warmup = F)
```


```{r out_mixmod1, echo=F, message=F, cache=T}

pars <- 
  c("theta", "theta_delta", "omega", "alpha", "p",
  "se[1]", "se[2]", "se[3]", "se[4]", "se[5]", "se[6]", "se[7]", "se[8]", 
  "se[9]", "se[10]","sp[1]", "sp[2]", "sp[3]", "sp[4]", "sp[5]", "sp[6]", 
  "sp[7]", "sp[8]", "sp[9]", "sp[10]")

out_mix1 <- matrix(0, length(all_surv), length(pars))
out_mix1_lb <- matrix(0, length(all_surv), length(pars))
out_mix1_ub <- matrix(0, length(all_surv), length(pars))
alpha_sd = rep(1, length(all_surv))
p_ub = rep(1, length(all_surv))
alpha_sd[9] <- 0.5
p_ub[9] <- 0.9

for(i in 1:length(all_surv)){
  
  print(i)
  y <- all_surv[[i]]
    
  fit <- sampling(
    mixmod1,  
    thin = 1,
    data = list(
         N = length(y), 
         y = y, 
         alpha_sd = alpha_sd[i],
         p_ub = p_ub[i]),
    chains = 3,            
    warmup = 1000,         
    iter = 4000,          
    cores = 3,
    seed = 111,
    refresh = 0
    )
  
  out_mix1[i, ] <- summary(fit)$summary[pars, 1]
  out_mix1_lb[i, ] <- summary(fit)$summary[pars, 4]
  out_mix1_ub[i, ] <- summary(fit)$summary[pars, 8]
  
}
  options(scipen=999)
  rownames(out_mix1) <- names(all_surv)
  colnames(out_mix1) <- pars
  rownames(out_mix1_lb) <- names(all_surv)
  colnames(out_mix1_lb) <- pars
  rownames(out_mix1_ub) <- names(all_surv)
  colnames(out_mix1_ub) <- pars
  
```


```{stan, output.var="senspmod", include=F, cache=T}

data {
  int N;
  int N_se;
  int N_sp;
  int y;
  int x;
  int z;
}

parameters {
  real<lower=0,upper=1> p;
  real<lower=0,upper=1> se;
  real<lower=0,upper=1> sp;
}

transformed parameters {
  real<lower=0,upper=1> p_obs;
  p_obs = se * p + (1 - sp) * (1 - p);
}

model {
   y ~ binomial(N, p_obs);
   x ~ binomial(N_se, se);
   z ~ binomial(N_sp, sp);
}

```


```{r out_senspmod, echo=F, message=F, warning=F, cache=T}

out_sensp <- matrix(0, length(all_surv), 3)

for(i in 1:length(all_surv)){
  
  y <- ifelse(all_surv[[i]] > 1, 1, 0) 
  
  fit <- sampling(
    senspmod,  
    thin = 1,
    data = list(
         y = sum(y),
         x = 166, 
         z = 901, 
         N = length(y), 
         N_se = 179, 
         N_sp = 910),
    chains = 3,            
    warmup = 1000,         
    iter = 4000,          
    cores = 3,
    seed = 111,
    refresh = 0           
  )

  out_sensp[i, ] <- summary(fit)$summary["p", c(1, 4, 8)]

}
  options(scipen=999)
  rownames(out_sensp) <-  names(all_surv)
  
```


```{stan, output.var="mixmod2", include=F, cache=T}

data {
  int N;
  vector[N] y;
  real nuInf_sd;
}

parameters {
  ordered[2] thetaInf; /// means for 2-component mixture in positives
  real<lower=0> delta; /// difference: mean infected - mean uninfected
  vector<lower=0>[2] nuInf; /// SDs for mixture in positives
  real<lower=0, upper=1> p;
  real<lower=0, upper=1> q;
}

transformed parameters{
  vector[3] theta;
  vector[3] nu;
  vector[3] pr;
  theta[1] = q * thetaInf[1] + (1 - q) * thetaInf[2] - delta;
  theta[2] = thetaInf[1];
  theta[3] = thetaInf[2];
  nu[1] = 0.42;
  nu[2] = nuInf[1];
  nu[3] = nuInf[2];
  pr[1] = 1 - p;
  pr[2] = p * q;
  pr[3] = p * (1 - q);
}

model {

  vector[3] log_pr= log(pr);

  // priors
  thetaInf ~ normal(0, 10);
  delta ~ normal(0, 1.83); 
  nuInf ~ lognormal(0.28, nuInf_sd); // median of lognormal = 1.32
  
  //likelihood
  for (n in 1:N) {
      vector[3] lps = log_pr; 
      for (k in 1:3){
        lps[k] += normal_lpdf(y[n] | theta[k], nu[k]);
      }
      target += log_sum_exp(lps);
    }
}

```


```{r out_mixmod2, echo=F, message=F, cache=T}

pars <- c("theta[1]", "theta[2]", "theta[3]", 
          "nu[2]", "nu[3]", "p", "q")

out_mix2 <- matrix(0, length(all_surv), length(pars))
out_mix2_lb <- matrix(0, length(all_surv), length(pars))
out_mix2_ub <- matrix(0, length(all_surv), length(pars))

for(i in 1:length(all_surv)){

print(i)  
nuInf_sd = rep(0.2, length(all_surv))
nuInf_sd[c(4, 5, 8, 9, 13, 15)] <- 0.05

  y <- all_surv[[i]]
    
  fit <- sampling(
    mixmod2,
    thin = 1,
    data = list(
         N = length(y),
         y = y,
         nuInf_sd = nuInf_sd[i]),
    chains = 3,
    warmup = 1000,
    iter = 4000,
    cores = 3,
    seed = 111,
    refresh = 0
  )
  
  out_mix2[i, ] <- summary(fit)$summary[pars, 1]
  out_mix2_lb[i, ] <- summary(fit)$summary[pars, 4]
  out_mix2_ub[i, ] <- summary(fit)$summary[pars, 8]
  
}
  options(scipen=999)
  rownames(out_mix2) <- names(all_surv)
  colnames(out_mix2) <- pars
  rownames(out_mix2_lb) <- names(all_surv)
  colnames(out_mix2_lb) <- pars
  rownames(out_mix2_ub) <- names(all_surv)
  colnames(out_mix2_ub) <- pars

```


```{r, eval=F, include=F}
 traceplot(fit, 
           pars = c("theta", "nu", "p", "q"), 
           inc_warmup = F)
```


**Test 1: pre-covid samples + PCR positives**

```{r test1, echo=F, message=F, cache=T}
test1 <- c(validation$`Pre COVID`, validation$`PCR+ asymp`, validation$`PCR+ symp`)
prop_pos <- (length(validation$`PCR+ asymp`) + length(validation$`PCR+ symp`)) / length(test1)
prop_pos

fit <- sampling(
  mixmod1,  
  thin = 1,
  data = list(
         N = length(test1), 
         y = test1, 
         nu = 0.42, 
         alpha_sd = 1,
         p_ub = 1),
  chains = 3,            
  warmup = 1000,         
  iter = 4000,          
  cores = 3,  
  seed = 111,
  refresh = 0           
)

options(scipen=999)
out_test1 <- summary(fit)$summary
out_test1[c("theta", "theta_delta", "omega", "alpha","p"), c(1, 4, 8)]

```


**Test 2: infected mixture**

```{r test2, echo=F, message=F, cache=T}

set.seed(116)
test_uninf <- rnorm(200, mean=0, sd=0.42)
test_inf1 <- rnorm(25, mean=2, sd=1.4)
test_inf2 <- rnorm(25, mean=0.5, sd=1.4)

test2 <- c(test_uninf, test_inf1, test_inf2)

fit <- sampling(
  mixmod1,  
  thin = 1,
  data = list(
         N = length(test2), 
         y = test2, 
         nu = 0.42, 
         alpha_sd = 1,
         p_ub = 1),
  chains = 3,            
  warmup = 1000,         
  iter = 4000,          
  cores = 3,  
  seed = 111,
  refresh = 0           
)

options(scipen=999)
out_test2 <- summary(fit)$summary
out_test2[c("theta", "theta_delta", "omega", "alpha","p"), c(1, 4, 8)]

```


**Test 3: uninfected mixture**

```{r test3, echo=F, message=F, cache=T}

set.seed(116)
test_uninf1 <- rnorm(100, mean=-0.3, sd=0.42)
test_uninf2 <- rnorm(100, mean=0.3, sd=0.42)
test_inf1 <- rnorm(50, mean=1.4, sd=1.4)

test3 <- c(test_uninf1, test_uninf2, test_inf1)

fit <- sampling(
  mixmod1,  
  thin = 1,
  data = list(
         N = length(test3), 
         y = test3, 
         nu = 0.42, 
         alpha_sd = 1,
         p_ub = 1),
  chains = 3,            
  warmup = 1000,         
  iter = 4000,          
  cores = 3,  
  seed = 111,
  refresh = 0           
)

options(scipen=999)
out_test3 <- summary(fit)$summary
out_test3[c("theta", "theta_delta", "omega", "alpha","p"), c(1, 4, 8)]

```


**Table 1: Mixture model parameter estimates**

```{r table 1, echo=F, include=T}

mean_un <- NULL
mean_un_ci <- NULL
mean_inf <- NULL
mean_inf_ci <- NULL
scale <- NULL
scale_ci <-NULL
skew <- NULL
skew_ci <- NULL
se <- NULL
se_ci <- NULL
sp <- NULL
sp_ci <- NULL

for(i in 1:length(all_surv)){
   
   mean_un[i] <- round(out_mix1[i, "theta"], digits = 2)
  
   mean_un_ci[i] <- paste(
     toString(round(out_mix1_lb[i, "theta"], digits = 2)), "," ,
     toString(round(out_mix1_ub[i, "theta"], digits = 2)))

   mean_inf[i] <- round(out_mix1[i, "theta_delta"], digits = 2)
  
   mean_inf_ci[i] <- paste(
     toString(round(out_mix1_lb[i, "theta_delta"], digits = 2)), "," ,
     toString(round(out_mix1_ub[i, "theta_delta"], digits = 2)))

   scale[i] <- round(out_mix1[i, "omega"], digits = 2)
  
   scale_ci[i] <- paste(
     toString(round(out_mix1_lb[i, "omega"], digits = 2)), "," ,
     toString(round(out_mix1_ub[i, "omega"], digits = 2)))

   skew[i] <- round(out_mix1[i, "alpha"], digits = 2)
  
   skew_ci[i] <- paste(
     toString(round(out_mix1_lb[i, "alpha"], digits = 2)), "," ,
     toString(round(out_mix1_ub[i, "alpha"], digits = 2)))
   
   se[i] <- round(out_mix1[i, "se[7]"], digits = 2)
  
   se_ci[i] <- paste(
     toString(round(out_mix1_lb[i, "se[7]"], digits = 2)), "," ,
     toString(round(out_mix1_ub[i, "se[7]"], digits = 2)))
   
   sp[i] <- round(out_mix1[i, "sp[7]"], digits = 2)
  
   sp_ci[i] <- paste(
     toString(round(out_mix1_lb[i, "sp[7]"], digits = 2)), "," ,
     toString(round(out_mix1_ub[i, "sp[7]"], digits = 2)))
   
   }   

 table_S1 <- cbind(mean_un, mean_un_ci, mean_inf, mean_inf_ci,
             scale, scale_ci, skew, skew_ci, se, se_ci, sp, sp_ci)
 
 rownames(table_S1) <- names(all_surv)
 
 print(table_S1, quote = F)

 write.csv(table_S1, file = "output/table_S1.csv")
 
```


**Table 2: Comparison of cumulative incidence estimates**

```{r table 2, echo = F, include=T}

p_crude <- NULL
popsize <- NULL
p_sensp <- NULL
p_sensp_lb <- NULL
p_sensp_ub <- NULL
p_sensp_ci <- NULL
p_mix1 <- NULL
p_mix1_lb <- NULL
p_mix1_ub <- NULL
p_mix1_ci <- NULL
p_mix2 <- NULL
p_mix2_lb <- NULL
p_mix2_ub <- NULL
p_mix2_ci <- NULL

for(i in 1:length(all_surv)){
   
   p_crude[i] <-  round(100 * sum(ifelse(all_surv[[i]] > 1, 1, 0)) 
                         / length(all_surv[[i]]), digits = 1)
   
   popsize[i] <- length(all_surv[[i]])
   
   p_sensp[i] <- round(100 * out_sensp[i, 1] , digits = 1)
  
   p_sensp_lb[i] <- round(100 * out_sensp[i, 2], digits = 1)
  
   p_sensp_ub[i] <- round(100 * out_sensp[i, 3], digits = 1)

   p_sensp_ci[i] <- paste(toString(p_sensp_lb[i]),"-",toString(p_sensp_ub[i]))
   
   p_mix1[i] <- round(100 * out_mix1[i, "p"], digits = 1)
  
   p_mix1_lb[i] <- round(100 * out_mix1_lb[i, "p"], digits = 1)
  
   p_mix1_ub[i] <- round(100 * out_mix1_ub[i, "p"], digits = 1)

   p_mix1_ci[i] <- paste(toString(p_mix1_lb[i]),"-",toString(p_mix1_ub[i]))
  
   p_mix2[i] <- round(100 * out_mix2[i, "p"], digits = 1)
  
   p_mix2_lb[i] <- round(100 * out_mix2_lb[i, "p"], digits = 1)
  
   p_mix2_ub[i] <- round(100 * out_mix2_ub[i, "p"], digits = 1)

   p_mix2_ci[i] <- paste(toString(p_mix2_lb[i]),"-",toString(p_mix2_ub[i]))
   
}

table_S2 <- cbind(popsize, p_crude, p_sensp, p_sensp_ci,
            p_mix1, p_mix1_ci, p_mix2, p_mix2_ci)

rownames(table_S2) <- names(all_surv) 

print(table_S2, quote = F)

write.csv(table_S2, file = "output/table_S2.csv")

```

**Prevalence ratio**

```{r}
pr = p_mix1/p_sensp
quantile(pr)
```


```{r}
mean(mean_un)
mean(mean_inf)
mean(scale)
mean(skew)
mean(se)
mean(sp)
```

**Weighted average of region-specific estimates**

sum(p_mix1[1:8]*popsize[1:8])/sum(popsize[1:8])

**Figure 1: Pre-COVID and PCR+ samples**

```{r fig_hist_pcr_precovid, echo = FALSE, message = FALSE, warning = FALSE}

yneg <- validation$`Pre COVID`

plotpre <-
  ggplot(data.frame(yneg), aes(x = yneg)) +
  geom_histogram(aes(y = ..density..), 
                 binwidth=0.1, fill = "white", color = "black") +
  theme_classic() +
  stat_function(fun = dnorm, 
                args = list(mean = mean(yneg), sd = sd(yneg))) +
  scale_x_continuous(
    breaks = c(-2, -1, 0, 1, 2, 3, 4, 5, 6), 
    labels = c("0.25", "0.5", "1", "2", "4", "8", "16", "32", "64"),
    limits = c(-2.1, 6.2)) +
  geom_vline(xintercept = 1, lty = "dashed") +
  xlab("") + 
  theme(text = element_text(size = 12)) +
  ggtitle("Pre-COVID-19")

ypos <- c(validation$`PCR+ asymp`, validation$`PCR+ symp`)

plotpcr <- 
  ggplot(data.frame(ypos), aes(x = ypos)) +
  geom_histogram(aes(y = ..density..), fill = "white", color = "black") +
  theme_classic() +
  stat_function(fun = dnorm, 
                args = list(mean = mean(ypos), sd = sd(ypos))) +
  scale_x_continuous(
    breaks = c(-2, -1, 0, 1, 2, 3, 4, 5, 6),
    labels = c("0.25", "0.5", "1", "2", "4", "8", "16", "32", "64"),
    limits = c(-2.1, 6.2)) +
  geom_vline(xintercept = 1, lty = "dashed") +
  xlab("OD ratio") + 
  theme(text = element_text(size = 12)) +
  ggtitle("PCR positive")

hist_pcr_precovid <-
  plot_grid(plotpre, plotpcr, ncol=1, rel_heights = c(1.5, 1))

ggsave("output/hist_pcr_precovid.pdf")

```

**Figure 2: Fitted distributions**


```{r fig_fitted_mix1, echo=F, message=F, warning=F}

lodr <- NULL
survey <- NULL

for(i in 1:length(all_surv)){
  lodr <- c(lodr, all_surv[[i]])
  survey <- c(survey, rep(i, length(all_surv[[i]])))
}

df1 <- data.frame(lodr, survey)
df1$surveynum <- df1$survey
df1$survey <- factor(df1$survey)
levels(df1$survey) <- names(all_surv) 

hist_kbts <- ggplot(df1[df1$surveynum <= 8, ], aes(x = lodr)) +
    geom_histogram(aes(y = ..density..), fill = "white", color = "black", size = 0.5) +
    theme_classic() +
    theme(text = element_text(size = 12)) +
    xlab("OD ratio") +
    ggtitle("Blood Donors") +
    theme(plot.title = element_text(size = 16, face = "bold")) +
    scale_x_continuous(breaks = c(-1, 0, 1, 2, 3, 4, 5),
                     labels = c("0.5", "1", "2", "4", "8", "16", "32"),
                     limits = c(-1.8, 5.2)) +
    scale_y_continuous(breaks = c(0, 0.5, 1, 2, 3, 4, 5),
                     limits = c(0, 1.3)) +
    facet_wrap(~survey) 

hist_other <- ggplot(df1[df1$surveynum > 8, ], aes(x = lodr)) +
    geom_histogram(aes(y = ..density..), fill = "white", color = "black") +
    theme_classic() +
    theme(text = element_text(size = 12)) +
    xlab("OD ratio") +
    ggtitle("Other surveys") +
    theme(plot.title = element_text(size = 16, face = "bold")) +
    scale_x_continuous(breaks = c(-1, 0, 1, 2, 3, 4, 5),
                     labels = c("0.5", "1", "2", "4", "8", "16", "32"),
                     limits = c(-1.8, 5.2)) +
    scale_y_continuous(breaks = c(0, 0.5, 1, 2, 3, 4, 5),
                     limits = c(0, 1.3)) +
    facet_wrap(~survey) 

denun <- function(x, theta, nu, p){
  (1-p) * dnorm(x = x, mean = theta, sd = nu)}

deninf <- function(x, theta_delta, omega, alpha, p){
  xi = theta_delta - sqrt(2/3.142) * omega * alpha/sqrt(1 + alpha^2)
  p * dsn(x = x, xi = xi, omega = omega, alpha = alpha)}

denmix <- function(x, theta, nu, theta_delta, omega, alpha, p){
  denun(x, theta, nu, p) + deninf(x, theta_delta, omega, alpha, p)}

addcurveun <- function(i) {
            geom_area(data = subset(df1, survey == levels(df1$survey)[i]), 
              stat = "function", fun = denun, fill = "#00AFBB", alpha = 0.4,
              args = list(theta = out_mix1[i, "theta"], 
                          nu = 0.42, 
                          p = out_mix1[i, "p"]))
  }

addcurveinf <- function(i) {   
            geom_area(data = subset(df1, survey == levels(df1$survey)[i]), 
              stat = "function", fun = deninf, fill = "#FC4E07", alpha = 0.4,
              args = list(theta_delta = out_mix1[i, "theta_delta"],
                          omega = out_mix1[i, "omega"],
                          alpha = out_mix1[i, "alpha"],
                          p = out_mix1[i, "p"]))
}

addcurvemix <-function(i) {
            stat_function(data = subset(df1, survey == levels(df1$survey)[i]), 
              fun = denmix, color = "black", lwd = 0.3,
              args = list(theta = out_mix1[i, "theta"], 
                          nu = 0.42,
                          theta_delta = out_mix1[i, "theta_delta"],
                          omega = out_mix1[i, "omega"],
                          alpha = out_mix1[i, "alpha"],
                          p = out_mix1[i, "p"])) 
}

pl_kbts <- hist_kbts

for(i in 1:8){  
  pl_kbts <- pl_kbts + addcurveun(i)
  pl_kbts <- pl_kbts + addcurveinf(i)
  pl_kbts <- pl_kbts + addcurvemix(i)
}

 pl_other <- hist_other
 
for(i in 9:length(all_surv)){  
  pl_other <- pl_other + addcurveun(i)
  pl_other <- pl_other + addcurveinf(i)
  pl_other <- pl_other + addcurvemix(i)
}

pl_all <- plot_grid(pl_kbts, pl_other)

ggsave("output/fig_fitted_mix1_all.pdf", pl_all, height = 9, width = 12)

```

**Figure 3: Comparison of estimates**

```{r fig_mix_vs_cutoff, include=T, echo=F}

pop_kbts <- 
  c(
    "Central, n=225",
    "Coast, n=435",
    "Eastern/N.Eastern, n=702",
    "Mombasa, n=802",
    "Nairobi, n=361",
    "Nyanza, n=584",
    "Rift Valley, n=508",
    "Western, n=106")

pop_anc <- 
  c(
    "Kilifi Oct, n=264",
    "Kilifi Dec, n=155",   
    "Nairobi, n=196"
  )

pop_hcw <- 
  c(
    "Busia, n=301",
    "Kilifi, n=200",   
    "Nairobi, n=183"
  )

pop_tru <-
  c(
    "Busia, n=365",
    "Magarini, n=101",
    "Malaba, n=364"
  )

p_all <- c(p_sensp, p_mix1)
p_all_lb <- c(p_sensp_lb, p_mix1_lb)
p_all_ub <- c(p_sensp_ub, p_mix1_ub)
method <- c(rep("threshold", length(all_surv)), rep("mixture", length(all_surv)))
y <- c(c(20:13, 11:9, 7:5, 3:1 ) - 0.1 , c(20:13, 11:9, 7:5, 3:1 ) + 0.1)
df2 <- data.frame(y, method, p_all, p_all_lb, p_all_ub) 

ggplot(data = df2, aes(x = p_all, y = y, color = method)) +
    geom_point() +
    geom_errorbarh(aes(xmin = p_all_lb, xmax = p_all_ub)) +
    scale_x_continuous(breaks = c(10, 20, 30, 40, 50, 60, 70, 80, 90)) +
    scale_y_continuous(
      breaks = 21:1,  
      labels = c(expression(bold("BLOOD DONORS")), pop_kbts, 
               expression(bold("ANC ATTENDEES")), pop_anc, 
               expression(bold("HEALTH CARE WORKERS")), pop_hcw, 
               expression(bold("TRUCK DRIVERS")), pop_tru)) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.grid.major.x = element_line( size=.1, color="black" )) +
    ylab("") +
    xlab("% previously infected") +
    scale_color_brewer(palette="Paired")

ggsave("output/fig_mix_vs_cut.pdf")

```


**Figure S1: Cumulative cases and timeline of serosurveys** 

```{r fig_timeline, include=T, echo=F}

df5 <- 
  world_covid %>%
  filter(countriesAndTerritories == "Kenya") %>%
  arrange(year, month, day) %>%
  mutate(cumCases = cumsum(cases)) %>%
  mutate(dateRep = dmy(dateRep)) 

ggplot(df5, aes(x = dateRep, y = cumCases)) +
  geom_area(color = "lightgray", fill = "lightgray") +
  scale_x_date(breaks = "1 month", date_labels = "%b") +
  theme_minimal(base_size = 12) +
  xlab("") +
  ylab("Total cases") +
  geom_segment(aes(x = dmy("30/7/2020"), y = 5000, xend = dmy("25/8/2020"), yend = 5000), lwd=1) +
  annotate("text", x = dmy("12/8/2020"), y = 8000, label = "ANC/HCW Nairobi") +
  geom_segment(aes(x = dmy("20/8/2020"), y = 15000, xend = dmy("30/9/2020"), yend = 15000), lwd=1) +
  annotate("text", x = dmy("10/9/2020"), y = 18000, label = "Blood donors") +
  geom_segment(aes(x = dmy("18/9/2020"), y = 25000, xend = dmy("23/10/2020"), yend = 25000), lwd=1) +
  annotate("text", x = dmy("4/10/2020"), y = 28000, label = "ANC Kilifi 1") +
  geom_segment(aes(x = dmy("30/9/2020"), y = 35000, xend = dmy("23/10/2020"), yend = 35000), lwd=1) +
  annotate("text", x = dmy("12/10/2020"), y = 38000, label = "TRU Magarini") +
  geom_segment(aes(x = dmy("13/10/2020"), y = 45000, xend = dmy("15/10/2020"), yend = 45000), lwd=1) +
  annotate("text", x = dmy("14/10/2020"), y = 48000, label = "TRU Malaba/Busia") +
  geom_segment(aes(x = dmy("13/10/2020"), y = 55000, xend = dmy("15/10/2020"), yend = 55000), lwd=1) +
  annotate("text", x = dmy("14/10/2020"), y = 58000, label = "TRU Malaba/Busia") +
  geom_segment(aes(x = dmy("19/10/2020"), y = 65000, xend = dmy("23/10/2020"), yend = 65000), lwd=1) +
  annotate("text", x = dmy("23/10/2020"), y = 68000, label = "HCW Busia") +
  geom_segment(aes(x = dmy("13/10/2020"), y = 75000, xend = dmy("4/12/2020"), yend = 75000), lwd=1) +
  annotate("text", x = dmy("7/11/2020"), y = 78000, label = "HCW Kilifi") +
  geom_segment(aes(x = dmy("24/10/2020"), y = 85000, xend = dmy("24/11/2020"), yend = 85000), lwd=1) +
  annotate("text", x=dmy("9/11/2020"), y=88000, label="ANC Kilifi 2")
  
 ggsave("output/timeline.pdf")
 
```


**Figure S2: ROC curves** 

```{r fig_roc, echo=F, message=F, warning=F}

cut_values <- c(-2, -1.5, -1, -0.5, 0, 0.5, 1, 1.5, 2, 5)

se <- 
  out_mix1 %>%
  data.frame() %>%
  rownames_to_column("survey") %>%
  select(c("survey", starts_with("se"))) %>%
  pivot_longer(cols = 2:11, names_to = "cut", values_to = "se") %>%
  mutate(cut = factor(cut, labels = cut_values)) %>%
  mutate(cut = as.numeric(as.vector(cut))) 

 sp <- 
  out_mix1 %>%
  data.frame() %>%
  rownames_to_column("survey") %>%
  select(c("survey", starts_with("sp"))) %>%
  pivot_longer(cols = 2:11, names_to = "cut", values_to = "sp") %>%
  mutate(cut = factor(cut, labels = cut_values)) %>%
  mutate(cut = as.numeric(as.vector(cut)))
 
df3 <- 
  se %>% 
  inner_join(sp, by = c("survey","cut")) %>%
  mutate(survey = factor(survey, levels = names(all_surv))) %>%
  mutate(surveynum = as.numeric(survey)) %>%
  mutate(fpr = 1 - sp)

ggplot(df3[df3$surveynum <= 8, ], aes(x = fpr , y = se)) +
    geom_line() +
    geom_abline(intercept = 0, slope = 1, lty = "dotted") +
    theme_classic() +
    theme(text = element_text(size = 10)) +
    xlab("1 - Specificity") +
    ylab("Sensitivity") +
    ggtitle("Blood Donors") +
    theme(plot.title = element_text(size = 11, face = "bold")) +
    facet_wrap(~survey) 

ggsave("output/roc_kbts.pdf")

ggplot(df3[df3$surveynum > 8, ], aes(x = fpr , y = se)) +
    geom_line() +
    geom_abline(intercept = 0, slope = 1, lty = "dotted") +
    theme_classic() +
    theme(text = element_text(size = 10)) +
    xlab("1 - Specificity") +
    ylab("Sensitivity") +
    ggtitle("Other Surveys") +
    theme(plot.title = element_text(size = 11, face = "bold")) +
    facet_wrap(~survey) 

ggsave("output/roc_other.pdf")

```

**Figure S3: Distributions predicted by the 3-components mixture model** 

```{r fig_fitted_mix2, echo=F, message=F, warning=F}

denun <- function(x, theta, nu, p){
  (1 - p) * dnorm(x = x, mean = theta, sd = nu)}

deninf1 <- function(x, theta, nu, p, q){
  p * q * dnorm(x = x, mean = theta, sd = nu) 
}

deninf2 <- function(x, theta, nu, p, q){
  p * (1 - q) * dnorm(x = x, mean = theta, sd = nu) 
}

denmix <- function(x, theta, nu, p, q){
  (1 - p) * dnorm(x = x, mean = theta[1], sd = nu[1]) +
  p * q * dnorm(x = x, mean = theta[2], sd = nu[2]) +
  p * (1 - q) * dnorm(x = x, mean = theta[3], sd = nu[3])     
}

addcurveun <- function(i) {
            geom_area(data = subset(df1, survey == levels(df1$survey)[i]), 
              stat = "function", fun = denun, fill = "#00AFBB", alpha = 0.4,
              args = list(theta = out_mix2[i, "theta[1]"], 
                          nu = 0.42, 
                          p = out_mix2[i, "p"]))
  }

addcurveinf1 <- function(i) {   
            geom_area(data = subset(df1, survey == levels(df1$survey)[i]), 
              stat = "function", fun = deninf1, fill = "#FC4E07", alpha = 0.4,
              args = list(theta = out_mix2[i, "theta[2]"],
                          nu = out_mix2[i, "nu[2]"],
                          p = out_mix2[i, "p"],
                          q = out_mix2[i, "q"]))
}

addcurveinf2 <- function(i) {   
            geom_area(data = subset(df1, survey == levels(df1$survey)[i]), 
              stat = "function", fun = deninf2, fill = "#FF0000", alpha = 0.6,
              args = list(theta = out_mix2[i, "theta[3]"],
                          nu = out_mix2[i, "nu[3]"],
                          p = out_mix2[i, "p"],
                          q = out_mix2[i, "q"]))
}

addcurvemix <-function(i) {
            stat_function(data = subset(df1, survey == levels(df1$survey)[i]), 
              fun = denmix, color = "black", lwd = 0.3,
               args = list(theta = c(out_mix2[i, "theta[1]"],
                                    out_mix2[i, "theta[2]"],
                                    out_mix2[i, "theta[3]"]),
                          nu = c(0.42,
                                 out_mix2[i, "nu[2]"],
                                 out_mix2[i, "nu[3]"]),
                          p = out_mix2[i, "p"],
                          q = out_mix2[i, "q"])
                         )
}

pl_kbts <- hist_kbts 

for(i in 1:8){  
  pl_kbts <- pl_kbts + addcurveun(i)
  pl_kbts <- pl_kbts + addcurveinf1(i)
  pl_kbts <- pl_kbts + addcurveinf2(i)
  pl_kbts <- pl_kbts + addcurvemix(i)
}

pl_other <- hist_other

for(i in 9:length(all_surv)){  
  pl_other <- pl_other + addcurveun(i)
  pl_other <- pl_other + addcurveinf1(i)
  pl_other <- pl_other + addcurveinf2(i)
  pl_other <- pl_other + addcurvemix(i)
}


pl_kbts

pl_other

ggsave("output/fig_fitted_mix2_kbts.pdf", pl_kbts)
ggsave("output/fig_fitted_mix2_other.pdf", pl_other)


```

**Figure S4: Comparison of estimates from skew normal and 3-component model** 

```{r fig_mix1_vs_mix2, include=T, echo=F}

p_all <- c(p_mix1, p_mix2)
p_all_lb <- c(p_mix1_lb, p_mix2_lb)
p_all_ub <- c(p_mix1_ub, p_mix2_ub)
model <- c(rep("skew normal", length(all_surv)), rep("3-component", length(all_surv)))
y <- c(c(20:13, 11:9, 7:5, 3:1 ) - 0.1 , c(20:13, 11:9, 7:5, 3:1 ) + 0.1)
df4 <- data.frame(y, model, p_all, p_all_lb, p_all_ub) 

ggplot(data = df4, aes(x = p_all, y = y, color = model)) +
    geom_point() +
    geom_errorbarh(aes(xmin = p_all_lb, xmax = p_all_ub)) +
    scale_x_continuous(breaks = c(10, 20, 30, 40, 50, 60, 70, 80, 90)) +
    scale_y_continuous(
      breaks = 21:1,  
      labels = c(expression(bold("BLOOD DONORS")), pop_kbts, 
               expression(bold("ANC ATTENDEES")), pop_anc, 
               expression(bold("HEALTH CARE WORKERS")), pop_hcw, 
               expression(bold("TRUCK DRIVERS")), pop_tru)) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.grid.major.x = element_line( size=.1, color="black" )) +
    ylab("") +
    xlab("% previously infected") +
    scale_color_brewer(palette="Paired")

ggsave("output/fig_mix1_vs_mix2.pdf")

```



